Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-10-01 vertical-thermal-review.Rmd
reworked abstract
@NidhiVinod , I reworked the abstract. It's now exactly at the 200 word limit. Please see what you think.

Text | word count | other | n
---|---|-----|--------------------
Total word count (excluding summary, references and legends) | 	8565 |	No. of figures | 	6 (all colour)
Summary|  	183 | 	No. of Tables | 	2  
Introduction|	495 |	No of Supporting Information files | 	6
Review of vertical gradients |	5976	||	 
Implications |	1753	 ||
Conclusions |	156	||
Acknowledgments | ##	||

@@

Rising temperatures are profoundly influencing forests on many scales with potentially strong variation vertically across forest strata. Using published research and new analyses, we evaluate how environmental conditions, leaf temperatures, and foliar traits and metabolism vary across vertical gradients, shaping tree ecology and ecosystem function. In forests with dense canopies, canopy leaves are exposed to higher solar radiation and evaporative demand than understory leaves, and can experience elevated leaf temperature ($T_{leaf}$), particularly when transpirational cooling is curtailed by limited stomatal conductance.  However, foliar traits that also vary strongly across height or light gradients can partially mitigate the elevation of $T_{leaf}$ in the upper canopy. Leaf metabolism generally increases with height across the vertical gradient, and tall trees have higher absolute metabolic capacity and growth at both individual and ecosystem levels. Yet, differences in thermal sensitivity and damage thresholds across the gradient are modest. Under climate warming and increased drought, larger trees can face disproportionate stress, yet understory trees have fewer cooling mechanisms and thus may be impacted the most under hot, humid conditions, or when the buffering provided by large trees is lost.

>>>>

Text | word count | other | n
---|---|-----|--------------------
Total word count (excluding summary, references and legends) | 	8565 |	No. of figures | 	6 (all colour)
Summary|  	200 | 	No. of Tables | 	2  
Introduction|	496 |	No of Supporting Information files | 	6
Review of vertical gradients |	5976	||	 
Implications |	1753	 ||
Conclusions |	156	||
Acknowledgments | ##	||

@@

Rising temperatures are influencing forests on many scales, with potentially strong variation vertically across forest strata. 
Using published research and new analyses, we evaluate how microclimate, leaf temperatures, and foliar traits and metabolism vary vertically in forests, shaping tree ecology and ecosystem function. 
In forests with dense canopies, upper-canopy leaves are exposed to the highest solar radiation and evaporative demand, which can elevate leaf temperature ($T_{leaf}$), particularly when transpirational cooling is curtailed by limited stomatal conductance.
However, foliar traits also vary across height or light gradients, partially mitigating the elevation of upper-canopy $T_{leaf}$.
Leaf metabolism generally increases with height across the vertical gradient, yet differences in thermal sensitivity and damage thresholds across the gradient appear modest.
Scaling up, tall trees have higher absolute metabolic capacity and growth at both individual and ecosystem levels, yet are disproportionately vulnerable to drought and damaging $T_{leaf}$'s, particularly under climate change. 
In contrast, understory trees benefit from a buffered microclimate but have fewer cooling mechanisms and thus may be disproportionately impacted under hot, humid conditions, or when the buffering provided by large trees is lost.
As the climate changes, integrating the patterns and mechanisms reviewed here into models will be critical to forecasting forest-climate feedbacks.
